Verification of a Deterministic Random Bits Generator

### 2015-12-31 paper/paper.tex
modified paper with suggestions from prof. appel

\section{Introduction}
A DRBG is a program that takes a source of randomness, or entropy, and stretches it out into a string of pseudorandom bits. \stdtitle{NIST SP 800-90A} \cite{nist} DRBG provides backtracking resistance and, optionally, prediction resistance. Backtracking resistance means that even if the whole (secret) state of the DRBG is known at time $t$, an adversary will not be able to distinguish between a truly random bitstring and a bitstring generated by the DRBG at some time before $t$. Prediction resistance is the complement, providing the same assurance for some time including and after $t$. Here, “time” is counted in terms of calls to the \lstinline{generate} function.

@@

\subsection{Machine-Checked Proofs}
I did the verification in Coq, an open-sourced proof assistant. The C code is compiled into its abstract syntax tree (AST) in Coq using the CompCert certified (machine-checked) compiler. I wrote functional specifications (functional programs) in Coq that closely match \stdtitle{NIST SP 800-90A}. Using the functional programs, I created API specifications using the VST framework. API specifications reason about how the AST deals with memory and local variables. I then proved the AST correct with respect to the API specifications.

@@

\section{A Functional Representation of Entropy} \label{entropy}
Before even understanding what a DRBG does, I need to explain how entropy is represented.

A DRBG needs some entropy input to act as a source of randomness. According to \stdtitle{NIST SP 800-90A}, the source of entropy input can be either an approved entropy source, a nondeterministic random bits generator (NRBG), or another DRBG. The entropy API used in NIST is the following:

@@

Note that in Coq, as well as many typed programming languages, an \lstinline{option} type is a type that can simply not exist (\lstinline{null} in the language of C and Java). It is effectively equivalent to this:

@@

I also defined a \lstinline{get_bits} function that, if possible, get the consecutive requested number of bits from the stream and returns a new stream that advances the index by the requested number of bits. If \lstinline{get_bits} encounters an error (a \lstinline{None} value), it returns a new stream that skips this error the next time it is called, symbolizing a movement forward in time resulting in more entropy being available.

@@

The distinction between a “working state” and a “state handle” is \stdtitle{NIST SP 800-90A} presents each of the function as first doing error checking then calling an inner algorithm. The error checking part is common to all types of DRBG, not just the HMAC-based one. The inner algorithm is specific to each type of DRBG, and the algorithm only needs the information stored in the working state (and the working state type does not need to be the same for each type of DRBG). See \stdtitle{NIST SP 800-90A \S 8.3} for more details.

Note that \lstinline{Z} is Coq’s representation of arbitrary-precision integers, and is used here to represent a byte. That is, the value of the integer must be between $0$ and $255$, inclusive. While I could have represented each byte as a tuple of a \lstinline{Z} and a proof that the value of that \lstinline{Z} is in the byte range, keeping that fact separate allowed me to write a functional specification that closely resembles \stdtitle{NIST SP 800-90A}. The fact that if the input state contains bytes, then the output will also contain bytes can easily be provided as separate lemmas about the individual functions.

\subsection{The Update Function}
Before showing any of the public function, I must first define a helper function, the \lstinline{update} function. See \stdtitle{NIST SP 800-90A \S 10.1.2.2} for the corresponding NIST specification. The \lstinline{update} function takes a \lstinline{key}, a \lstinline{value}, and an optional nonce and returns a new \lstinline{key} and a new \lstinline{value}. This function will be called at least once in each of the public functions, and the goal is to provide backtracking resistance. This is achieved by having the secret internal state of the DRBG (the \lstinline{key}) be the output of HMAC, an irreversible function.

@@

The \lstinline{instantiate} function initializes the DRBG state with randomness from a source of entropy. The caller is responsible for specifying the minimum security strength for this DRBG, which for an HMAC-based DRBG translates to “how many bits of entropy should be requested?”. In addition to entropy, the DRBG also asks for a nonce, which is roughly defined as a string of bits with at least \lstinline{1/2 security_strength} bits of entropy. In fact, this nonce can even be taken form the same source of entropy. For the purposes of this project, I am choosing to specialize the specification so that the nonce is \textit{always} taken from the same source of entropy. This is compatible with \stdtitle{mbedTLS}.

@@

Note that this is just the “outer” function that does error checking on the parameters, as mentioned earlier. This outer function is purely generic; different types of DRBG can share this piece of code. The actual algorithm for computing an initial working state is left as a dependency in the first parameter. All parameters before \lstinline{entropy_stream} are implementation specific constants. For example, \lstinline{prediction_resistance_supported} is a flag denoting whether or not this particular implementation can provide prediction resistance.

@@

As with \lstinline{instantiate}, the parameters up to but excluding \lstinline{entropy_stream} are implementation specific constants.

Note that both the \lstinline{reseed} and \lstinline{generate} functions allow users to pass in an additional sequence of bytes. Users do not need to provide this (and implementations can choose to omit this feature), but it is meant to provide any additional bits of entropy or to provide personalization of this specific call.

@@

Note that \lstinline{Fixpoint} is a recursive function where at least one parameter is structurally getting smaller, allowing Coq to figure out the proof of termination automatically. In this case, it is the last parameter, the natural number \lstinline{count}, which is initialized with \lstinline{1}, meaning that this recursive function is only allowed to go one level deep.

@@

The \lstinline{generate} algorithm first checks if the number of times \lstinline{generate} has been called\\ (\lstinline{reseed_counter}) is too high, and if so, returns a flag saying a reseed is needed. Otherwise, it repeatedly HMACs the current \lstinline{value} with the current \lstinline{key}, append the result to the output, and store the result as the new \lstinline{value}. This is why the \lstinline{value} part of the DRBG state is not considered secret, and is really the output of the DRBG. The output is then trimmed down to the requested number of bytes, and \lstinline{reseed_counter} is incremented by \lstinline{1}.

\section{Verification}
\subsection{HMAC Interface}\label{hmac_interface}
In this section, I explain the (pre-existing) interface of HMAC, of which the DRBG specification is a client of.

@@

The state machine representation of HMAC was created for this project. \cite{hmac} describes the previous verification effort on HMAC, but it represented HMAC in a more low-level format. See \S \ref{oldhmac} for more details.

@@

\begin{enumerate}
\item The field \lstinline{entropy_len} in the C struct is equivalent to \lstinline{security_strength} defined in the functional state.
\item The field \lstinline{reseed_interval} are stored explicitly in the C struct but are treated as a global constant in NIST SP 800-90A and so in the functional specification.
\end{enumerate}

@@

The concrete state is required to be a tuple of this type by the VST framework; I just gave it a convenient name. Here, \lstinline{mdstate} is the byte value of the struct representing the HMAC context. \lstinline{val} is used to represent the bytes representation of the fields in the C struct.

@@

Here, \lstinline{length} and \lstinline{buffer} are Coq (logical) variables, while \lstinline{_buffer_length} and \lstinline{_buffer} are predefined identifiers. Note that identifiers are just unique positive integers.

@@

Note that line counts include empty lines as well as comments. The compilation time was measured on a Macbook Pro with Intel Core i7, 2.5 GHz, 16 gigabytes of memory.

@@

One of the goals of my project is to serve as a bleeding-edge adopter of the VST framework and the HMAC API specification. As is the nature of bleeding-edge adoption, I encountered many difficulties, such as a proof tactic failing silently and Verifiable C's theory of pointer comparisons being too weak. These difficulties are fixed as they get reported back to the developers of VST. In this section, I highlight one particular difficulty with the HMAC API specification.

@@

Note that the \stdtitle{mbedTLS} message digest interface declares the following functions:

@@

The next \lstinline{WITH} value is the pointer value, \lstinline{ctx}, pointing to the DRBG struct. The next two \lstinline{WITH} values are the abstract and concrete DRBG states, respectively. Skipping the next \lstinline{WITH} value, \lstinline{kv}, for the moment, the last \lstinline{WITH} value is the content of the struct denoting what kind of HMAC is being used, \lstinline{info_contents}. In the \lstinline{PROP} clause, the sequence \lstinline{V} part of the abstract state is enforced to be the size of the output of the HMAC function and enforced to contain valid bytes. This is an invariant that the update function and all public functions enforce, and so will be seen throughout all API specifications. In the \lstinline{SEP} clause, \lstinline{ctx} is said to point to an area in memory that contains the concrete state, and the concrete state is related to the abstract state by the equivalence definition. The info pointer from the concrete state also points to a valid address containing \lstinline{info_contents}. In the post condition, there is a new abstract state calculated by \lstinline{hmac256drbgabs_hmac_drbg_update}, which is just a wrapper around the functional specification with the HMAC256 function used. In the \lstinline{SEP} part, there exists a corresponding concrete state pointed to by \lstinline{ctx} that is equivalent to the new abstract state. This new concrete state’s info pointer also points to the same valid memory location containing \lstinline{info_contents}. This is what the definition \lstinline{hmac256drbgabs_common_mpreds} encapsulates.

Lastly, the \lstinline{WITH} value \lstinline{kv} represents the pointer value to the SHA K vector, which is a global constant array in the SHA-256 implementation. For more information, see the verification of SHA-256 \cite{sha}.

@@

Notice a new notation here: \lstinline{EX ret_value:_}. This can be read literally as “there exists a variable \lstinline{ret_value} that satisfies the postcondition”. This existential variable captures the return value of the function, and its value is enforced in the \lstinline{PROP} part of the postcondition. The return value is enforced to be 0 if the reseed succeeds, a specific non-zero value if the reseed fails because of a failure to get entropy, and any non-zero value if the reseed fails for any other reason.

Note that \lstinline{mbedtls_HMAC256_DRBG_reseed_function} is just a wrapper around the functional specification of the reseed function, with the reseed algorithm and other dependencies filled in. \lstinline{hmac256drbgabs_reseed} returns the resulting abstract state from the reseed function if the reseed succeeds, or the initial abstract state otherwise.

@@

Again, much like the \lstinline{update} function, the next three \lstinline{WITH} values are the pointer value to the DRBG state, the abstract DRBG state, and the concrete DRBG state. The only new property is in the \lstinline{PROP} part of the precondition, requiring the \lstinline{entropy_len} component to be 32. Note that in \stdtitle{NIST SP 800-90A} and therefore in the functional specification, this is a global, fixed constant, but \stdtitle{mbedTLS} varies this value based on the type of HMAC used and fixes the value upon instantiation. With HMAC256, this is fixed to be 32.

>>>>

\section{Introduction}
A deterministic random bits generator (DRBG) is a program that takes a source of randomness, or entropy, and stretches it out into a much longer string of pseudorandom bits. \stdtitle{NIST SP 800-90A} \cite{nist} DRBG provides backtracking resistance and, optionally, prediction resistance. Backtracking resistance means that even if the whole (secret) state of the DRBG is known at time $t$, an adversary will not be able to distinguish between a truly random bitstring and a bitstring generated by the DRBG at some time before $t$. Prediction resistance is the complement, providing the same assurance for some time including and after $t$. Here, “time” is counted in terms of calls to the \lstinline{generate} function.

@@

\subsection{Machine-Checked Proofs}
I did the verification in Coq, an open-sourced proof assistant. The C code is compiled into its abstract syntax tree (AST) in Coq using the CompCert certified (proved correct with a machine-checked proof) compiler. I wrote functional specifications (functional programs) in Coq that closely match \stdtitle{NIST SP 800-90A}. Using the functional programs, I created API specifications using the VST framework. API specifications reason about how the AST deals with memory and local variables. I then proved the AST correct with respect to the API specifications.

@@

\section{A Functional Representation of Entropy} \label{entropy}
A DRBG needs some entropy input to act as a source of randomness. According to \stdtitle{NIST SP 800-90A}, the source of entropy input can be either an approved entropy source, a nondeterministic random bits generator (NRBG), or another DRBG. The entropy API used in NIST is the following:

@@

In Coq, as well as many typed programming languages, an \lstinline{option} type is a type whose members can either exist and have a value or not exist. Types in C and Java are implicitly optional, with the members that do not exist being represented by \lstinline{null}. It is effectively equivalent to this:

@@

The name \lstinline{catastrophic_error} came from \stdtitle{NIST SP 800-90A} and is used to denote errors that are not user errors. User errors include providing an optional nonce that is too long and requesting prediction resistance when the DRBG state has not been initialized with prediction resistance capability. An example of catastrophic errors is the entropy source failing to provide the requested amount of entropy.

I also defined a \lstinline{get_bits} function that, if possible, gets the consecutive requested number of bits from the stream and returns a new stream that advances the index by the requested number of bits. If \lstinline{get_bits} encounters an error (a \lstinline{None} value), it returns a new stream that skips this error the next time it is called, symbolizing a movement forward in time resulting in more entropy being available.

@@

The distinction between a “working state” and a “state handle” is that \stdtitle{NIST SP 800-90A} presents each of the functions as first doing error checking then calling an inner algorithm. The error checking part is common to all types of DRBG, not just the HMAC-based one. The inner algorithm is specific to each type of DRBG, and the algorithm only needs the information stored in the working state (and the working state type does not need to be the same for each type of DRBG). See \stdtitle{NIST SP 800-90A \S 8.3} for more details.

\lstinline{Z} is Coq’s representation of arbitrary-precision integers, and is used here to represent a byte. That is, the value of the integer must be between $0$ and $255$, inclusive. While I could have represented each byte as a tuple of a \lstinline{Z} and a proof that the value of that \lstinline{Z} is in the byte range, keeping that fact separate allowed me to write a functional specification that closely resembles \stdtitle{NIST SP 800-90A}. The fact that if the input state contains bytes, then the output will also contain bytes can easily be provided as separate lemmas about the individual functions.

\subsection{The Update Function}
Before showing any of the public functions, I must first define a helper function, \lstinline{update}. See \stdtitle{NIST SP 800-90A \S 10.1.2.2} for the corresponding NIST specification. The \lstinline{update} function takes a \lstinline{key}, a \lstinline{value}, and an optional nonce and returns a new \lstinline{key} and a new \lstinline{value}. This function will be called at least once in each of the public functions, and the goal is to provide backtracking resistance. This is achieved by having the secret internal state of the DRBG (the \lstinline{key}) be the output of HMAC, an irreversible function.

@@

The \lstinline{instantiate} function initializes the DRBG state with randomness from a source of entropy. The caller is responsible for specifying the minimum security strength for this DRBG, which for an HMAC-based DRBG translates to “how many bits of entropy should be requested?”. In addition to entropy, the DRBG also asks for a nonce, which is roughly defined as a string of bits with at least \lstinline{1/2 security_strength} bits of entropy. In fact, this nonce can even be taken from the same source of entropy. For the purposes of this project, I am choosing to specialize the specification so that the nonce is \textit{always} taken from the same source of entropy. This is compatible with \stdtitle{mbedTLS}.

@@

This is just the “outer” function that does error checking on the parameters, as mentioned earlier. This outer function is purely generic; different types of DRBG can share this piece of code. The actual algorithm for computing an initial working state is left as a dependency in the first parameter. All parameters before \lstinline{entropy_stream} are implementation-specific constants. For example, \lstinline{prediction_resistance_supported} is a flag denoting whether or not this particular implementation can provide prediction resistance.

@@

As with \lstinline{instantiate}, the parameters up to but excluding \lstinline{entropy_stream} are implementation-specific constants.

Both the \lstinline{reseed} and \lstinline{generate} functions allow users to pass in an additional sequence of bytes. Users do not need to provide this (and implementations can choose to omit this feature), but it is meant to provide any additional bits of entropy or to provide personalization of this specific call.

@@

\lstinline{Fixpoint} is a recursive function where at least one parameter is structurally getting smaller, allowing Coq to figure out the proof of termination automatically. In this case, it is the last parameter, the natural number \lstinline{count}, which is initialized with \lstinline{1}, meaning that this recursive function is only allowed to go one level deep.

@@

The \lstinline{generate} algorithm first checks if the number of times \lstinline{generate} has been called\\ (\lstinline{reseed_counter}) is too high, and if so, returns a flag saying a reseed is needed. Otherwise, it repeatedly HMACs the current \lstinline{value} with the current \lstinline{key}, appends the result to the output, and stores the result as the new \lstinline{value}. This is why the \lstinline{value} part of the DRBG state is not considered secret, and is really the output of the DRBG. The output is then trimmed down to the requested number of bytes, and \lstinline{reseed_counter} is incremented by \lstinline{1}.

\section{Verification}
\subsection{HMAC Interface}\label{hmac_interface}
In this section, I explain the (pre-existing) interface of HMAC, of which the DRBG specification is a client.

@@

The state machine representation of HMAC was created for this project. Beringer \textit{et al.} \cite{hmac} describe the previous verification effort on HMAC, but they represented HMAC in a more low-level format. See \S \ref{oldhmac} for more details.

@@

\begin{enumerate}
\item The field \lstinline{entropy_len} in the C struct is equivalent to \lstinline{security_strength} defined in the functional state.
\item The field \lstinline{reseed_interval} is stored explicitly in the C struct but is treated as a global constant in NIST SP 800-90A and so in the functional specification.
\end{enumerate}

@@

The concrete state is required to be a tuple of this type (to match the fields of the C struct) by the VST framework; I just gave it a convenient name. Here, \lstinline{mdstate} is the byte value of the struct representing the HMAC context. \lstinline{val} is used to represent the bytes representation of the fields in the C struct.

@@

Here, \lstinline{length} and \lstinline{buffer} are Coq (logical) variables, while \lstinline{_buffer_length} and \lstinline{_buffer} are the names of local variables (function parameters) of the C program. C program identifiers are represented in VST as unique positive integers.

@@

Line counts here include empty lines as well as comments. The compilation time was measured on a Macbook Pro with Intel Core i7, 2.5 GHz, 16 gigabytes of memory.

@@

One of the goals of my project is to serve as a bleeding-edge adopter of the VST framework and the HMAC API specification. As is the nature of bleeding-edge adoption, I encountered many difficulties, such as proof tactics failing silently and Verifiable C's theory of pointer comparisons being too weak. These difficulties were fixed as they got reported back to the developers of VST. In this section, I highlight one particular difficulty with the HMAC API specification.

@@

The \stdtitle{mbedTLS} message digest interface declares the following functions:

@@

The next \lstinline{WITH} value is the pointer value, \lstinline{ctx}, pointing to the DRBG struct. The next two \lstinline{WITH} values are the abstract and concrete DRBG states, respectively. The last \lstinline{WITH} value is the content of the struct denoting what kind of HMAC is being used, \lstinline{info_contents}. In the \lstinline{PROP} clause, the sequence \lstinline{V} part of the abstract state is enforced to be the size of the output of the HMAC function and enforced to contain valid bytes. This is an invariant that the update function and all public functions enforce, and so will be seen throughout all API specifications. In the \lstinline{SEP} clause, \lstinline{ctx} is said to point to an area in memory that contains the concrete state, and the concrete state is related to the abstract state by the equivalence definition. The info pointer from the concrete state also points to a valid address containing \lstinline{info_contents}. In the post condition, there is a new abstract state calculated by \lstinline{hmac256drbgabs_hmac_drbg_update}, which is just a wrapper around the functional specification with the HMAC256 function used. In the \lstinline{SEP} part, there exists a corresponding concrete state pointed to by \lstinline{ctx} that is equivalent to the new abstract state. This new concrete state’s info pointer also points to the same valid memory location containing \lstinline{info_contents}. This is what the definition \lstinline{hmac256drbgabs_common_mpreds} encapsulates.

The \lstinline{WITH} value \lstinline{kv} represents the pointer value to the SHA K vector, which is a global constant array in the SHA-256 implementation. For more information, see the verification of SHA-256 \cite{sha}.

@@

Notice a new notation here: \lstinline{EX ret_value:_}. This can be read literally as “there exists a variable \lstinline{ret_value} that satisfies the postcondition”. This existential variable captures the return value of the function, and its value is enforced in the \lstinline{PROP} part of the postcondition. The return value is enforced to be 0 if the reseed succeeds, a specific nonzero value if the reseed fails because of a failure to get entropy, and any nonzero value if the reseed fails for any other reason.

\lstinline{mbedtls_HMAC256_DRBG_reseed_function} is just a wrapper around the functional specification of the reseed function, with the reseed algorithm and other dependencies filled in. \lstinline{hmac256drbgabs_reseed} returns the resulting abstract state from the reseed function if the reseed succeeds, or the initial abstract state otherwise.

@@

Again, much like the \lstinline{update} function, the next three \lstinline{WITH} values are the pointer value to the DRBG state, the abstract DRBG state, and the concrete DRBG state. The only new property is in the \lstinline{PROP} part of the precondition, requiring the \lstinline{entropy_len} component to be 32. In \stdtitle{NIST SP 800-90A} and therefore in the functional specification, this is a global, fixed constant, but \stdtitle{mbedTLS} varies this value based on the type of HMAC used and fixes the value upon instantiation. With HMAC256, this is fixed to be 32.
